---
title: "How to estimate the growth rate stratified by groups?"
format: 
  html:
    code-link: true
editor: source
editor_options: 
  chunk_output_type: console
date: last-modified
toc: true
toc_float: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Ingredients

- Calculate the daily growth rate stratified by groups.
- Use the `covidregionaldataUK` data set from the `{covidregionaldata}` R package.
- Use the "date of report" in `cases_new` column of `covidregionaldataUK` to calculate the case counts.
- Keep the first 4 months in `covidregionaldataUK` as period of analysis.

## Steps in code

```{r}
#| warning: false

# Calculate the daily growth rate of an outbreak stratified by groups

# Load required packages
library(tidyverse)
library(covidregionaldata)
library(incidence2)
library(i2extras)

# Load covid outbreak data
raw_dat <- 
  covidregionaldataUK %>% 
  as_tibble() %>% 
  filter(magrittr::is_in(region,c("England", "Scotland", 
                                  "Northern Ireland", "Wales"))) %>% 
  filter(date > lubridate::ymd(20200701)) %>%
  filter(date < lubridate::ymd(20201101))

# Read covid outbreak data
raw_dat

# Plot the raw data
raw_dat %>%
  ggplot() +
  aes(x = date, y = cases_new) +
  geom_line() +
  facet_wrap(vars(region))

# Convert the data to an incidence object stratified by groups
dat_group <- 
  incidence2::incidence(
    x = raw_dat, 
    counts = "cases_new",
    date_index = "date",
    interval = "day",
    groups = "region"
  )

dat_group

# Model the incidence
out_group <- 
  i2extras::fit_curve(
    dat_group, 
    model = "poisson", 
    alpha = 0.05
  )

# Print the model result
i2extras::growth_rate(out_group)

# Plot the model result with prediction intervals
plot(out_group, ci = FALSE, pi = TRUE)

# Read plot documentation for an incidence2_fit object
# ?plot.incidence2_fit
```

## Steps in detail

- (pending)

## Related

- [Explanation on incidence, growth rate, doubling time, and Poisson model](https://www.reconlearn.org/slides/outbreak-statistics/outbreak-statistics.html#28)
