---
title: "How to get an incidence object from days to weeks?"
format: 
  html:
    code-link: true
editor: source
editor_options: 
  chunk_output_type: console
date: last-modified
toc: true
toc_float: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Ingredients

- Get weekly case counts from daily collected data.
- Use the `linelist` object in the `ebola_sim_clean` list from the `{outbreaks}` R package. `linelist` is an outbreak data set. 
- Use the `date_of _onset` column of `linelist` to calculate the case counts. 

## Steps in code

```{r}
#| warning: false
#| fig-height: 4
#| fig-width: 10

# Get the weekly growth rate of an outbreak

# Load required packages
library(outbreaks)
library(incidence2)
library(dplyr)
library(patchwork)

# Load the simulated Ebola outbreak data
data(ebola_sim_clean)

# Extract the first element of the list
linelist <- ebola_sim_clean$linelist

# Read data collected per day
as_tibble(linelist)

# Get an incidence2 object with case counts per day
# from data collected per day
incidence2_day <- 
  incidence2::incidence(
    x = linelist, 
    date_index = "date_of_onset",
    interval = "day"
  )

# Read case counts per day
incidence2_day

# Plot incidence2 object with case counts per day
plot_day <- plot(incidence2_day)

# Get an incidence2 object with case counts per week
# from data collected per day
incidence2_week <- 
  incidence2::incidence(
    x = linelist, 
    date_index = "date_of_onset",
    interval = "week"
  )

# Read case counts per week
incidence2_week

# Plot incidence2 object with case counts per week
plot_week <- plot(incidence2_week)

# Add plots together
plot_day + plot_week
```

## Steps in detail

- The `outbreaks` package is loaded to access the simulated Ebola outbreak data.
- The `ebola_sim_clean` object from the package contains the simulated outbreak data.
- The `linelist` object contains the first list element from `ebola_sim_clean`.
- The `incidence()` function from the `incidence2` package converts the data object `linelist` to an `incidence2` class object.
- The `date_of_onset` column is used in the `date_index` argument as the onset dates of the outbreak.
- The `"day"` text string is used in the `interval` argument to count number of cases per day interval.
- The `"week"` text string is used in the `interval` argument to count number of cases per week interval.  
- The `incidence2` object provides observations arranged in descendant order with respect to the `date_index`.

Please note that the code assumes the necessary packages are already installed. If they are not, you can install them using first the `install.packages("pak")` function and then the `pak::pak()` function for both packages in CRAN or GitHub before loading them with `library()`. 

Additionally, make sure to adjust the serial interval distribution parameters according to the specific outbreak you are analyzing.

## Related

- [Explanation on incidence](https://www.reconlearn.org/slides/outbreak-statistics/outbreak-statistics.html#28)
