---
title: "How to extract the serial interval distribution from a systematic review?"
format: 
  html:
    code-link: true
editor: source
editor_options: 
  chunk_output_type: console
date: last-modified
toc: true
toc_float: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Ingredients

- Use the distribution estimated from the WHO Ebola Response Team in 2015.

## Steps in code

```{r}
#| warning: false

# Load required packages
library(epiparameter)
library(tidyverse)

# Access to the library of parameters as an epiparam object
# epidist_all <- epiparameter::epidist_db

# Filter by disease, epi_distribution, region, and year
epidist_ebola <- epiparameter::epidist_db(
    disease = "Ebola",
    epi_dist = "serial_interval",
    single_epiparameter = TRUE
    )

epidist_ebola

# Convert a row to an epidist class object
# epidist_convert <- as_epidist(epiparam_ebola)
# 
# epidist_convert

# Plot the epidist object
plot(epidist_ebola,day_range = 0:50)
```

## Steps in detail

- The `epiparameter` package is loaded to access the library of epidemiological parameters.
- The `epidist_db()` function extract a parameter by specifying the disease name in the `disease` argument, epidemiological distribution in the `epi_dist` argument, and author name in the `author` argument.
- The `epidist_db()` function creates an `epidist` object.
- The `plot` function creates a plot from `epidist` objects. 
<!-- - The `epiparam()` functions provide access to a data frame with all the names of diseases, epidemiological distribution, and authors. -->

Please note that the code assumes the necessary packages are already installed. If they are not, you can install them using first the `install.packages("pak")` function and then the `pak::pak()` function for both packages in CRAN or GitHub before loading them with `library()`. 

Additionally, make sure to adjust the serial interval distribution parameters according to the specific outbreak you are analyzing.

## Related

- [Explanation on epidemiological distributions](https://www.reconlearn.org/slides/outbreak-statistics/outbreak-statistics.html#14)
