---
title: "How to estimate the growth rate stratified by groups?"
format: 
  html:
    code-link: true
editor: source
editor_options: 
  chunk_output_type: console
date: last-modified
toc: true
toc_float: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Ingredients

- Calculate the daily growth rate stratified by groups.
- Use the `covidregionaldataUK` data set from the `{incidence2}` R package.
- Use the "date of report" in `cases_new` column of `covidregionaldataUK` to calculate the case counts.
- Keep the first 4 months in `covidregionaldataUK` as period of analysis.

## Steps in code

```{r}
#| warning: false

# Calculate the daily growth rate of an outbreak stratified by groups

# Load required packages
library(tidyverse)
library(incidence2)

# Load covid outbreak data
raw_dat <- 
  covidregionaldataUK %>% 
  as_tibble() %>% 
  filter(magrittr::is_in(region,c("England", "Scotland", 
                                  "Northern Ireland", "Wales"))) %>% 
  filter(date > lubridate::ymd(20200701)) %>%
  filter(date < lubridate::ymd(20201101))

# Read covid outbreak data
raw_dat

# Plot the raw data
# raw_dat %>%
#   ggplot() +
#   aes(x = date, y = cases_new) +
#   geom_line() +
#   facet_wrap(vars(region))

# Convert the data to an incidence object stratified by groups
dat_group <- 
  incidence2::incidence(
    x = raw_dat, 
    counts = "cases_new",
    date_index = "date",
    interval = "day",
    groups = "region"
  )

dat_group

# Model the incidence
fitted <- 
  dat_group %>% 
  nest(.key = "data") %>% 
    mutate(
        model  = lapply(
            data,
            function(x) glm(count ~ date_index, data = x, family = poisson)
        )
    )

# Print the model result
intervals <-
    fitted %>%
    mutate(result = Map(
        function(data, model) {
            data %>%
                ciTools::add_ci(
                    model,
                    alpha = 0.05,
                    names = c("lower_ci", "upper_ci")
                ) %>%
                as_tibble()
        },
        data,
        model
    )) %>%
    select(region, result) %>%
    unnest(result)

# Plot the model result with prediction intervals
plot(dat_group, angle = 45) +
    ggplot2::geom_line(
        ggplot2::aes(date_index, y = pred),
        data = intervals,
        inherit.aes = FALSE
    ) +
    ggplot2::geom_ribbon(
        ggplot2::aes(date_index, ymin = lower_ci, ymax = upper_ci),
        alpha = 0.2,
        data = intervals,
        inherit.aes = FALSE,
        fill = "#BBB67E"
    )

# Read plot documentation for an incidence2_fit object
# ?plot.incidence2_fit
```

## Steps in detail

- (pending)

## Related

- [Explanation on incidence, growth rate, doubling time, and Poisson model](https://www.reconlearn.org/slides/outbreak-statistics/outbreak-statistics.html#28)
