---
title: "How to estimate R and relative susceptibility from age-stratified pre- and post-epidemic serological data?"
format: 
  html:
    code-link: true
editor: source
editor_options: 
  chunk_output_type: console
date: last-modified
toc: true
toc_float: true
author: 
  - name: "Adam Kucharski"
    orcid: "0000-0001-8814-9421"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Ingredients

- Social mixing data from `socialmixr`
- Serological data measuring rise in influenza titres during the 2009 A/H1N1p pandemic in different age groups (i.e. proxy for infection)
- Want to estimate the reproduction number and relative susceptibility in older age groups from the pattern of infection, based on the final epidemic size in an age-stratified SIR-like model, following the methods of [Kucharski et al (2014)](https://journals.plos.org/plospathogens/article?id=10.1371/journal.ppat.1004206) and using Markov chain Monte Carlo.


## Steps in code

```{r}
#| warning: false

# Load required packages
library(finalsize)
library(socialmixr)
library(MCMCpack)
library(readr)
library(dplyr)
library(ggplot2)

# Load serological data
# source:
# Kwok KO, Cowling BJ, Wei VW, Wu KM, Read JM, Lessler J, Cummings DA, Peiris JS, Riley S. 
# Social contacts and the locations in which they occur as risk factors for influenza infection. 
# Proc Biol Sci. 2014 Aug 22;281(1789):20140709. doi: 10.1098/rspb.2014.0709. 
# PMID: 25009062; PMCID: PMC4100506.
# https://pmc.ncbi.nlm.nih.gov/articles/instance/4100506/
hk_serology <- readr::read_csv("https://github.com/epiverse-trace/howto/raw/refs/heads/main/data/rspb20140709supp2.csv")

# Allocate ages to bands
age_limits <- c(3,18,25,40,65)
age_group <- sapply(hk_serology$age,function(x){sum(x>=age_limits)})
hk_serology$age_group <- age_group

# Load HK social mixing data
contact_data <- socialmixr::contact_matrix(
  polymod,
  countries = "United Kingdom",
  age.limits = age_limits,
  symmetric = TRUE
)

demography_vector <- contact_data$demography$population

# Define model
model_function <- function(param){
  r0 <- param[1]
  alpha_val <- param[2]
  
  contact_matrix = t(contact_data$matrix)
  contact_matrix = contact_matrix / max(eigen(contact_matrix)$values)
  
  names(demography_vector) = contact_data$demography$age.group
  n_demo_grps = length(demography_vector)
  
  contact_matrix = contact_matrix / demography_vector
  
  susceptibility = matrix(
    data = c(1, alpha_val*rep(1,n_demo_grps-1)),
    n_demo_grps
  )
  
  n_risk_grps = 1L
  p_susceptibility = matrix(1, n_demo_grps, n_risk_grps)
  
  simulation_output <- finalsize::final_size(
    r0 = r0,
    contact_matrix = contact_matrix,
    demography_vector = demography_vector,
    susceptibility = susceptibility,
    p_susceptibility = p_susceptibility,
    solver = "iterative"
  )
  
  return(simulation_output)
}

# Define likelihood function
likelihood_function <- function(param,data) {
  # Ensure positive parameters
  if (any(param <= 0)) return(-Inf)

  simulation_output <- model_function(param)
  
  # Assuming non-informative priors for alpha and beta
  log_prior <- 0 # Can refine if have different priors
  
  # Calculate log-likelihood
  test_positive <- hk_serology$ffold
  log_likelihood <- log(
    test_positive*simulation_output$p_infected[hk_serology$age_group] + 
      (1-test_positive)*(1-simulation_output$p_infected[hk_serology$age_group])
  )
  log_likelihood_total <- sum(log_likelihood)
  
  # Check for valid output
  if(is.na(log_likelihood_total)){
    log_likelihood_total <- -Inf
  }
  
  # Return log-posterior (log-likelihood + log-prior)
  return(log_likelihood_total + log_prior)
}

## Set up MCMCpack for Monte Carlo estimation
initial_param <- c(r0=1.5, alpha=0.8) # Initial guess for shape and rate

n_mcmc <- 1000

output <- MCMCpack::MCMCmetrop1R(
  fun = likelihood_function, 
  theta.init = initial_param,
  mcmc = n_mcmc, # Number of MCMC iterations
  burnin = 100, # Burn-in period
  verbose = FALSE, # Turn off verbose output
  data = hk_serology
)

# Plot outputs
data_summary <- stats::aggregate(
  x = ffold ~ age_group,
  data = hk_serology, 
  FUN = function(x){mean(x == 1)}
)

sample_outputs <- output[sample(1:n_mcmc,10),]
sample_simulate <- apply(sample_outputs,1,model_function)

# Combine all list elements into a single data frame
combined_data <- do.call(rbind, lapply(1:length(sample_simulate), function(i) {
  cbind(sample_simulate[[i]], Simulation = i)
}))

combined_data$age_index <- base::match(
  x = combined_data$demo_grp,
  table = contact_data$demography$age.group
)

# Calculate mean and standard deviation of p_infected for each demo_grp
summary_data <- combined_data %>%
  dplyr::group_by(age_index) %>%
  dplyr::summarise(
    Mean_p_infected = mean(p_infected),
    SD_p_infected = sd(p_infected)
  ) %>%
  dplyr::ungroup()

# Plotting
summary_data %>% 
  dplyr::left_join(data_summary, by = c("age_index" = "age_group")) %>% 
  ggplot() +
  geom_point(aes(x = age_index, y = Mean_p_infected)) +
  geom_point(aes(x = age_index, y = ffold),pch=19,col="red") +
  ylim(0,1)
```

## Steps in detail
- We use serological data from the 2009 pandemic in Hong Kong in [Kwok et al (2009)](https://royalsocietypublishing.org/doi/10.1098/rspb.2014.0709), and social mixing data from POLYMOD in the UK as an assumed contact matrix using `{socialmixr}`.
- We define the likelihood of observing a given serological pattern by simulating age-specific final epidemic size using the `{finalsize}` package.
- We use [{MCMCpack}](https://cran.r-project.org/web/packages/MCMCpack/index.html) to estimate the parameters using MCMC.
- Finally, we plot the observed outputs against the model fits.
