---
title: "How to quantify the time-varying reproduction number (R~t~)?"
format: 
  html:
    code-link: true
editor: source
editor_options: 
  chunk_output_type: console
date: last-modified
toc: true
toc_float: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Ingredients

- Simulated outbreak data of Ebola Virus Disease that matches some key properties of the West African Ebola outbreak of 2014-2015 from the package `{outbreaks}`.
- The `{linelist}` package to keep tagged and validated columns in a line list data set.
- The `{incidence2}` package to generate aggregated incidence data with the daily number of reported cases.
- The `{epiparameter}` package to access to the serial interval estimated by the WHO Ebola Response Team in 2015.
- Assume that the serial interval distribution approximates the generation time.
- The `{EpiEstim}` package to estimate the time-varying reproduction number.

## Steps in code

```{r}
#| warning: false

# Load required packages
library(outbreaks)
library(incidence)
library(epiparameter)
library(EpiEstim)
library(tidyverse)

# Load the simulated Ebola outbreak data
data(ebola_sim_clean)

# Extract the first element of the list
linelist <- ebola_sim_clean$linelist

# Convert the data to an incidence object
incidence_data <- incidence::incidence(linelist$date_of_onset)

# Extract parameter by disease, distribution, author
epidist_ebola <- 
  epiparameter::epidist_db(
    disease = "Ebola",
    epi_name = "serial_interval",
    single_epiparameter = TRUE
    )

# Estimate the time-varying reproduction number
epiestim_output <- estimate_R(
  incid = incidence_data, 
  method = "parametric_si",
  config = make_config(
    list(
      mean_si = epidist_ebola$summary_stats$mean,
      std_si = epidist_ebola$summary_stats$sd
    )
  )
)

# Plot the time-varying reproduction number
plot(epiestim_output)
```

## Steps in detail

- The `outbreaks` package is loaded to access the simulated Ebola outbreak data.
- The `epiparameter` package is loaded to access the library of epidemiological parameters.

- The `ebola_sim_clean` object from the package contains the simulated outbreak data.
- The `linelist` object contains the first list element from `ebola_sim_clean`.
- The `incidence()` function from the `incidence` package converts the vector `date_of_onset` from the `linelist` data frame to an `incidence` class object.

- The `epidist_db()` function from the `epiparameter` package extract a parameter by specifying the disease name in the `disease` argument, epidemiological distribution in the `epi_name` argument, and author name in the `author` argument.

- The `estimate_R()` function from the `EpiEstim` package estimates the time-varying reproduction number (Rt). We provide the `incidence_data`, specify the method as `"parametric_si"` (parametric with a known serial interval), and pass the serial interval distribution parameters using the `make_config` function.
- The `plot` function creates three plots from the `estimate_R` class object.

Please note that the code assumes the necessary packages are already installed. If they are not, you can install them using first the `install.packages("pak")` function and then the `pak::pak()` function for both packages in CRAN or GitHub before loading them with `library()`. 

Additionally, make sure to adjust the serial interval distribution parameters according to the specific outbreak you are analyzing.

## Related

- [Explanation on the time-varying effective reproductive number](https://mrc-ide.github.io/EpiEstim/articles/full_EpiEstim_vignette.html)
