---
title: "How to estimate R and k from cluster size data using Markov chain Monte Carlo?"
format: 
  html:
    code-link: true
editor: source
editor_options: 
  chunk_output_type: console
date: last-modified
toc: true
toc_float: true
---

## Ingredients

- Use Bayesian estimation methods to estimate the reproduction number ($R$) and extent of superspreading, represented by the dispersion of a negative binomial distribution for individual-level seconday cases ($k$), from data on MERS-CoV- outbreak clusters.

- We will use [Markov chain Monte Carlo (MCMC)](https://en.wikipedia.org/wiki/Markov_chain_Monte_Carlo), specifically a simple Metropolis-Hastings algorithm to estimate the parameters.


## Steps in code

```{r}
#| warning: false

# Load required packages
library(epichains)
library(MCMCpack)
```


## Data
We use MERS cluster sizes from [Cauchemez et al, Lancet Inf Dis, 2013](https://www.thelancet.com/journals/laninf/article/PIIS1473-3099(13)70304-9/fulltext).
```{r}
# Define data
mers_clusters = c(rep(1,27),c(2,2),c(3,4),c(4,3),c(5,2),7,13,26)

# Show summary table of frequencies
freq_df <- as.data.frame(table(mers_clusters)); names(freq_df) <- c("Cluster size", "Frequency")

# Create a table for the HTML document
knitr::kable(freq_df, caption = "Frequencies of MERS Clusters")
```

## Likelihood
We define the likelihood of observing a specific cluster size distribution, given $R$ and $k$
```{r}
# Define likelihood function
lik_function <- function(param) {
  if (any(param <= 0)) return(-Inf) # Ensure positive parameters
  
  # Extract values of R and k
  r_val <- as.numeric(param[1])
  k_val <- as.numeric(param[2])

  # Define likelihood
  log_likelihood <- likelihood(
    chains = mers_clusters,
    statistic = "size",
    offspring_dist = rnbinom,
    size = k_val,
    mu = r_val
  )
  
  # Assume non-informative priors for R and k
  log_prior <- 0 # But could add informative priors here if required

  # Return log-posterior (log-likelihood + log-prior)
  return(log_likelihood + log_prior)
}
```

## Define MCMC parameters

```{r common-params}
# Define number of MCMC iterations
n_iter <- 1e4

# Define 'burn in' period for fitting, to be discarded
n_burn <- 1e3

# Initial guess for c(R,k):
init_param <- c(R=0.5, k=0.5)
```

## MCMCpack

[{MCMCpack}](https://cran.r-project.org/web/packages/MCMCpack/index.html) is an R package for Bayesian statistical inference through Markov Chain Monte Carlo (MCMC) methods, offering a broad array of algorithms and models for efficient and straightforward Bayesian estimation.

```{r MCMCpack-run, message=FALSE, warning=FALSE, fig.width=7, fig.height=6}
# Run MCMC to estimate parameters
result_mcmcpack <- MCMCmetrop1R(lik_function, 
                                theta.init = init_param, 
                                burnin = n_burn, 
                                mcmc = n_iter, 
                                thin = 1)

# Calculate effective sample size (i.e. measure of MCMC mixing)
ess_mcmcpack <- effectiveSize(result_mcmcpack)

# Plot posterior estimates
plot(result_mcmcpack)
```

Next, output a parameter table:
```{r}
# Define helper function to calculate median and 95% credible interval from data.frame of MCMC samples
get_param <- function(x){
  apply(x,2,function(y){val = signif(quantile(y,c(0.5,0.025,0.0975)),3);
                        val_text <- paste0(val[1]," (95%: CrI: ",val[2],"-",val[3],")")})
}

# Get posterior median and 95% CrI
posterior_estimates <- get_param(result_mcmcpack)

# Compile table
results_table <- data.frame(
  Package = "MCMCpack",
  Posterior_R = posterior_estimates[1],
  Posterior_k = posterior_estimates[2],
  ESS_R = ess_mcmcpack[1],
  ESS_k = ess_mcmcpack[2]
)

# Output the table with kable
knitr::kable(results_table, caption = "MCMC Comparison Table", align = 'c')
```

## Steps in detail
- The `epichains` package is loaded for the likelihood functions and `MCMCpack` for MCMC fitting.
- The likelihood is defined using the `likelihood()` function in epichains, with a negative binomial used (`rnbinom`).
- The MCMC is run using `MCMCmetrop1R()`, with number of iterations `mcmc` and burn in period `burnin` specified.

