---
title: "How to use _epidemics_ classes with _odin_-generated models"
format: 
  html:
    code-link: true
editor: source
editor_options: 
  chunk_output_type: console
date: last-modified
toc: true
toc_float: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Ingredients

This guide shows how to use some of the convenient features of _epidemics_, especially the classes that describe populations, interventions, and vaccination regimes, with models that are generated by _odin_.

This guide shows this interoperability by implementing the default model from _epidemics_ in _odin_.

```{r}
library(epidemics)
library(socialmixr)
library(odin)
library(data.table)
library(ggplot2)
```

## Steps in code

### Prepare population initial conditions

```{r}
# get social contacts data
polymod <- socialmixr::polymod
contact_data <- socialmixr::contact_matrix(
  polymod,
  countries = "United Kingdom",
  age.limits = c(0, 20, 40),
  symmetric = TRUE
)

# assume arbitrary populations for each demo group
demography_vector <- contact_data$demography$population

# prepare contact matrix, divide by leading eigenvalue and rowwise by popsize
contact_matrix <- t(contact_data[["matrix"]])
contact_matrix <- (contact_matrix / max(Re(eigen(contact_matrix)$values))) /
  demography_vector

# an intervention to close schools
close_schools <- intervention(
  type = "contacts",
  time_begin = 200,
  time_end = 260,
  reduction = matrix(c(0.5, 0.01, 0.01))
)

# view the intervention
close_schools

# initial conditions defined as proportions and converted to absolute values
initial_conditions <- matrix(
  c(S = 1 - 1e-6, E = 0, I = 1e-6, R = 0),
  nrow = length(demography_vector),
  ncol = 4L, byrow = TRUE
) * demography_vector

# an intervention to close schools
close_schools <- intervention(
  type = "contacts",
  time_begin = 200,
  time_end = 260,
  reduction = matrix(c(0.15,0.15,0.15))
)

# view the intervention
close_schools
```

### Define an epidemic model in _odin_

This is an SIR model that omits the 'exposed' compartment for this demonstration.

```{r}
# define SIR model
sir <- odin::odin({
  # NOTE: crude time-dependence of transmission rate
  # similar to a `<rate_intervention>`
  beta_reduce[] <- if (t > intervention_interval[1] && t < intervention_interval[2]) (1.0 - reduction[i]) else (1 + 0*reduction[i])

  # number of age groups taken from contact matrix passed by user
  n <- user()

  # FOI is contacts * infectious * transmission rate
  # returns a matrix, must be converted to a vector/1D array
  lambda_prod[, ] <- C[i, j] * I[j] * beta * beta_reduce[j]* beta_reduce[i]
  #lambda_prod2[, ] <- if (i != j) lambda_prod[i, j] * beta_reduce[i] else lambda_prod[i, j] # if want to keep diagonal the same
  lambda[] <- sum(lambda_prod[i, ])

  ## Derivatives
  deriv(S[]) <- -S[i] * lambda[i]
  deriv(E[]) <- (S[i] * lambda[i]) - (nu * E[i])
  deriv(I[]) <- (nu * E[i]) - (gamma * I[i])
  deriv(R[]) <- gamma * I[i]

  ## Initial conditions: passed by user
  initial(S[]) <- init_S[i]
  initial(E[]) <- init_E[i]
  initial(I[]) <- init_I[i]
  initial(R[]) <- init_R[i]

  ## parameters
  beta <- user()
  nu <- 1 / 3
  gamma <- 1 / 7
  C[, ] <- user() # user defined contact matrix
  init_S[] <- user()
  init_E[] <- user()
  init_I[] <- user()
  init_R[] <- user()
  reduction[] <- user()
  intervention_interval[] <- user()

  # dimensions - all rely on contact matrix
  dim(lambda_prod) <- c(n, n)
  #dim(lambda_prod2) <- c(n, n)
  dim(lambda) <- n
  dim(S) <- n
  dim(E) <- n
  dim(I) <- n
  dim(R) <- n
  dim(init_S) <- n
  dim(init_E) <- n
  dim(init_I) <- n
  dim(init_R) <- n
  dim(reduction) <- n
  dim(beta_reduce) <- n
  dim(C) <- c(n, n)
  dim(intervention_interval) <- 2
})
```

```{r}
# initialise model and run over time 0 - 600
mod <- sir$new(
  beta = 1.5 / 7,
  reduction = close_schools$reduction[,],
  intervention_interval = c(close_schools$time_begin, close_schools$time_end),
  C = contact_matrix, n = nrow(contact_matrix),
  init_S = initial_conditions[, 1],
  init_E = initial_conditions[, 2],
  init_I = initial_conditions[, 3],
  init_R = initial_conditions[, 4]
)
t <- seq(0, 600)
y <- mod$run(t)

# convert to data.table and plot infectious in each age class
y <- as.data.table(y)
y <- melt(y, id.vars = c("t"))

ggplot(y[variable %like% "I"]) +
  geom_vline(
    xintercept = c(close_schools[["time_begin"]], close_schools[["time_end"]]),
    colour = "red",
    linetype = "dashed",
    linewidth = 0.2
  ) +
  annotate(
    geom = "text",
    label = "Schools closed",
    colour = "red",
    x = 230, y = 400e3,
    angle = 90,
    vjust = "outward"
  ) +
  geom_line(
    aes(t, value, col = variable) 
  ) +
  scale_colour_brewer(
    palette = "Dark2",
    labels = rownames(contact_matrix),
    name = "Age group"
  ) +
  scale_y_continuous(
    labels = scales::comma,
    name = "Individuals infected"
  ) +
  labs(
    x = "Model time (days)"
  ) +
  theme_bw() +
  theme(
    legend.position = "top"
  )
```
