---
title: "Simulate Epidemic Attack Rates with Heterogeneous Social Contacts in the UK"
format: 
  html:
    code-link: true
editor: source
editor_options: 
  chunk_output_type: console
date: last-modified
toc: true
toc_float: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## What do we have?

- Basic reproduction number.
- Social contact survey and demography data of Country in the [POLYMOD](https://journals.plos.org/plosmedicine/article?id=10.1371/journal.pmed.0050074) study or [Social contact data](https://zenodo.org/communities/social_contact_data/records) repository.
- Susceptibility groups.
- Probability of infection in demographic and susceptibility groups.


## Steps in code

```{r}
#| warning: false

# Load packages
library(finalsize)
library(socialmixr)
library(tidyverse)


# Simple quick calculation with homogenous mixing -------------------------------------------
r0_input <- 2
finalsize::final_size(r0 = r0_input)

# Set up the transmission model -------------------------------------------

# load contact and population data from socialmixr::polymod
polymod <- socialmixr::polymod
contact_data <- socialmixr::contact_matrix(
  polymod,
  countries = "United Kingdom",
  age.limits = c(0, 5, 18, 40, 65),
  symmetric = TRUE
)

# prepare contact matrix and demography vector for use in model
contact_matrix <- t(contact_data$matrix) # transpose so R0 calculated correctly inside model
demography_vector <- contact_data$demography$population
names(demography_vector) <- rownames(contact_matrix)

# scale the contact matrix so the largest eigenvalue is 1.0
# this is to ensure that the overall epidemic dynamics correctly reflect
# the assumed value of R0
contact_matrix <- contact_matrix / max(Re(eigen(contact_matrix)$values))

# divide each row of the contact matrix by the corresponding demography
# this reflects the assumption that each individual in group {j} make contacts
# at random with individuals in group {i}
contact_matrix <- contact_matrix / demography_vector

n_demo_grps <- length(demography_vector)

# all individuals are equally and highly susceptible
n_susc_groups <- 1L
susc_guess <- 1.0

susc_uniform <- matrix(
  data = susc_guess,
  nrow = n_demo_grps,
  ncol = n_susc_groups
)

# Final size calculations also need to know the proportion of each demographic group {ð‘–} 
# that falls into the susceptibility group {ð‘—}. This distribution of age groups into 
# susceptibility groups can be represented by the demography-susceptibility distribution matrix.
p_susc_uniform <- matrix(
  data = 1.0,
  nrow = n_demo_grps,
  ncol = n_susc_groups
)


output <- finalsize::final_size(
  r0 = r0_input,
  contact_matrix = contact_matrix,
  demography_vector = demography_vector,
  susceptibility = susc_uniform,
  p_susceptibility = p_susc_uniform
)

output

output %>% 
  mutate(demo_grp = as_factor(demo_grp)) %>% 
  ggplot(aes(x = demo_grp, y = p_infected)) +
  geom_col() +
  ylim(0,1) +
  labs(
    x = "Age group",
    y = "Proportion infected",
    title = "Final size of an SIR epidemic",
    subtitle = "Fully susceptible population"
  )
```


## Steps in detail

<!-- OPTIONAL -->

<!-- reduce length of strings with a large language model like chatgpt -->

- This assume equal probability of infection in demographic and susceptibility groups. 
    - You can vary the probability of infection across demographic and susceptibility groups following the [Guide to construct susceptibility matrices](https://epiverse-trace.github.io/finalsize/articles/susceptibility_matrices.html)

<!-- must: keep to warn users on how to install packages -->

Please note that the code assumes the necessary packages are already installed. If they are not, you can install them using first the `install.packages("pak")` function and then the `pak::pak()` function for both packages in CRAN or GitHub before loading them with `library()`. 

<!-- optional: erase if no assumed distribution is needed -->


## Related

- [Explanation on Modelling heterogeneous social contacts](https://epiverse-trace.github.io/finalsize/articles/varying_contacts.html)
- [Explanation on Guide to constructing susceptibility matrices](https://epiverse-trace.github.io/finalsize/articles/susceptibility_matrices.html)
