---
title: "How to (active verb) ... ?"
format: 
  html:
    code-link: true
editor: source
editor_options: 
  chunk_output_type: console
date: last-modified
toc: true
toc_float: true
---

## Ingredients

- Nouns

## Steps in code

```{r}
#| warning: false

# Load required packages
library(tidyverse)

# Define the immunization effect
immunization_effect <- 0.25

# Read the immunization effect
immunization_effect
```

```{r}
# optionally, to read data
# from 
# https://github.com/epiverse-trace/cleanepi/blob/main/inst/extdata/test_df.RDS
# right click the "View Raw" link, 
# copy and paste that link below
data_path <- "https://github.com/epiverse-trace/cleanepi/raw/main/inst/extdata/test_df.RDS"

readr::read_rds(data_path)
```


## Steps in detail

<!-- OPTIONAL -->

<!-- reduce length of strings with a large language model like chatgpt -->

- `tidyverse` package is loaded to manage data frame objects.

<!-- must: keep to warn users on how to install packages -->

Please note that the code assumes the necessary packages are already installed. If they are not, you can install them using first the `install.packages("pak")` function and then the `pak::pak()` function for both packages in CRAN or GitHub before loading them with `library()`. 

<!-- optional: erase if no assumed distribution is needed -->

Additionally, make sure to adjust the serial interval distribution parameters according to the specific outbreak you are analyzing.

## Related

- [Explanation on topic](link)
